\textbf{SPECFEM3D Cartesian}\\ \textbf{User Manual}

### 2016-06-16 doc/USER_MANUAL/03_mesh_generation.tex
added notes about attenuation
updates doc

In each example subdirectory you will find a \texttt{README} file,
which explains in a step-by-step tutorial the workflow for the example.
Please feel free to contribute your own example to this package by
contacting the CIG Computational Seismology Mailing List \urlwithparentheses{cig-seismo@geodynamics.org}.

In some cases, to re-create the meshes for the examples given, just type "claro ./create\_mesh.py" or similar
from the command line ("claro" is the command to run CUBIT from the command line).

IMPORTANT note: in order to correctly set up GEOCUBIT and run the examples, please read the file called "EXAMPLES/README";
in particular, please make sure you correctly set up the Python paths as indicated in that file.

@@

Once the geometric model volumes in CUBIT are meshed, you prepare
the model for exportation with the definition of material blocks and
boundary surfaces. Thus, prior to exporting the mesh, you need to
define blocks specifying the materials and absorbing boundaries in
CUBIT. This process could be done automatically using the script \texttt{run\_cubit2specfem3d.py}
if the mesh meets some conditions or manually, following the block
convention:
\begin{description}
\item [{material\_name}] Each material should have a specific block defined
by a unique name. The name convention of the material is to start
with either 'elastic' or 'acoustic'. It must be then followed by a
unique identifier, e.g. 'elastic 1', 'elastic 2', etc. The additional
attributes to the block define the material description.

For an elastic material:
\begin{description}
\item [{material\_id}] An integer value which is unique for this material.
\item [{Vp}] P-wave speed of the material (given in m/s).
\item [{Vs}] S-wave speed of the material (given in m/s).
\item [{rho}] density of the material (given in kg/m$^{3}$).
\item [{Q}] quality factor to use in case of a simulation with attenuation
turned on. It should be between 1 and 9000. In case no attenuation
information is available, it can be set to zero. You can either specify a single Q value,
in which case it will be assumed to be pure shear attenuation $Q_{\mu}$, or two
separate values for bulk and shear attenuation, $Q_{\kappa}$ and $Q_{\mu}$ respectively.

@@

When you have an existing CUBIT (or similar) mesh stored in SPECFEM3D
format, i.e., if you have existing \textquotedbl{}nodes\_coords\_file\textquotedbl{}
and \textquotedbl{}mesh\_file\textquotedbl{} files but do not know
how to assign CPML flags to them, we have created a small serial Fortran
program that will do that automatically for you, i.e., which will
create the \textquotedbl{}absorbing\_cpml\_file\textquotedbl{} for
you. That program is utils/CPML/convert\_external\_layers\_of\_a\_given\_mesh\_to\_CPML\_layers.f90,
and a small Makefile is provided in that directory (utils/CPML).

IMPORTANT: it is your responsibility to make sure that in the input
CUBIT (or similar) mesh that this code will read in SPECFEM3D format
from files \textquotedbl{}nodes\_coords\_file\textquotedbl{} and \textquotedbl{}mesh\_file\textquotedbl{}
you have created layers of elements that constitute a layer of constant
thickness aligned with the coordinate grid axes (X, Y and/or Z), so
that this code can assign CPML flags to them. This code does NOT check
that (because it cannot, in any easy way). The mesh inside these CPML
layers does not need to be structured nor regular, any non-structured
mesh is fine as long as it has flat PML inner and outer faces, parallel
to the axes, and thus of a constant thickness. The thickness can be
different for the X, Y and Z sides. But for X it must not vary, for
Y it must not vary, and for Z it must not vary. If you do not know
the exact thickness, you can use a slightly LARGER value in this code
(say 2\% to 5\% more) and this code will fix that and will adjust
it; never use a SMALLER value otherwise this code will miss some CPML
elements.

@@

where \texttt{\textbf{domain\_ID}}\textbf{ is 1 for acoustic and 2
for elastic or viscoelastic materials,} \texttt{material\_ID} a unique
identifier, \texttt{rho} the density in $kg\, m^{-3}$, \texttt{vp}
the P-wave speed in $m\, s^{-1}$, \texttt{vs} the S-wave speed in
$m\, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag}
an identifier for anisotropic models. Note that the \texttt{Qkappa}
value is ignored by the code unless \texttt{FULL\_ATTENUATION\_SOLID}
is set. Note also that both \texttt{Qkappa} and \texttt{Qmu} are ignored
by the code unless \texttt{ATTENUATION} is set. If you want a model
with no \texttt{Qmu} attenuation, both set \texttt{ATTENUATION} to
\texttt{.false.} in the \texttt{Par\_file} and set \texttt{Qmu} to
9999 here. If you want a model with no \texttt{Qkappa} attenuation,
both set \texttt{FULL\_ATTENUATION\_SOLID} to \texttt{.false.} in
the \texttt{Par\_file} and set \texttt{Qkappa} to 9999 here. %%magnoni

>>>>

In each example subdirectory you will find a \texttt{README} file,
which explains in a step-by-step tutorial the workflow for the example.
Please feel free to contribute your own example to this package by
contacting the CIG Computational Seismology Mailing List \urlwithparentheses{cig-seismo@geodynamics.org}.\\

In some cases, to re-create the meshes for the examples given, just type
\begin{verbatim}
claro ./create_mesh.py
\end{verbatim}
or similar from the command line (\texttt{claro} is the command to run CUBIT from the command line).

{\it IMPORTANT:} In order to correctly set up GEOCUBIT and run the examples, please read the file called \texttt{EXAMPLES/README};
in particular, please make sure you correctly set up the Python paths as indicated in that file.

Concerning the script \texttt{create\_mesh.py}, you may find the option \texttt{use\_explicit} which chooses how to assign the material properties for the volume (or domain):
\begin{itemize}
\item[(1)] one way is to explicitly assign block attributes to the block of the corresponding volume, by commands like:
\begin{verbatim}
cubit.cmd('block '+str(id_block)+' attribute index 2 2800') # vp
\end{verbatim}
This is done when using \texttt{use\_explicit = 1} in the script. The final command:
\begin{verbatim}
cubit2specfem3d.export2SPECFEM3D('MESH/')
\end{verbatim}
will then create the corresponding \texttt{MESH/nummaterial\_velocity\_file} for all such defined block volumes/domains.

\item[(2)] the other option with \texttt{use\_explicit = 0} is to let GEOCUBIT deal with a dummy entry and then overwrite the \texttt{nummaterial\_velocity\_file} at the end with a corresponding command section:
\begin{verbatim}
f = open(nummaterial_velocity_file,'w')
\end{verbatim}
This second way is chosen by default because GEOCUBIT can handle partitions for several processes and glues everything together automatically. Thus, it adds some more sophistication when the model gets more complicated than just this single volume example.
\end{itemize}
You will find out by experimenting what is easier for your case.

@@

Once the geometric model volumes in CUBIT are meshed, you prepare
the model for exportation with the definition of material blocks and
boundary surfaces. Thus, prior to exporting the mesh, you need to
define blocks specifying the materials and absorbing boundaries in
CUBIT. This process could be done automatically using the script \texttt{run\_cubit2specfem3d.py}
if the mesh meets some conditions or manually, following the block
convention:
\begin{description}
\item [{material\_name}] Each material should have a specific block defined
by a unique name. The name convention of the material is to start
with either \textbf{'elastic'} or \textbf{'acoustic'}. It must be then followed by a
unique identifier, e.g. \textbf{'elastic 1'}, \textbf{'elastic 2'}, etc. The additional
attributes to the block define the material description.

For an elastic material:
\begin{description}
\item [{material\_id}] An integer value which is unique for this material.
\item [{Vp}] P-wave speed of the material (given in m/s).
\item [{Vs}] S-wave speed of the material (given in m/s).
\item [{rho}] density of the material (given in kg/m$^{3}$).
\item [{Q}] quality factor to use in case of a simulation with attenuation
turned on. It should be between 1 and 9000. In case no attenuation
information is available, it can be set to zero. You can either specify a single Q value,
in which case it will be assumed to be pure shear attenuation $Q_{\mu}$, or two
separate values for bulk and shear attenuation, $Q_{\kappa}$ and $Q_{\mu}$ respectively.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see doc/note\_on\_Qkappa\_versus\_Qp.pdf and utils/attenuation/conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90.

@@

When you have an existing CUBIT (or similar) mesh stored in SPECFEM3D
format, i.e., if you have existing \texttt{nodes\_coords\_file}
and \texttt{mesh\_file} files but do not know
how to assign CPML flags to them, we have created a small serial Fortran
program that will do that automatically for you, i.e., which will
create the \texttt{absorbing\_cpml\_file} for
you. That program is: \\
\texttt{utils/CPML/convert\_external\_layers\_of\_a\_given\_mesh\_to\_CPML\_layers.f90},\\
and a small Makefile is provided in that directory (utils/CPML).

IMPORTANT: it is your responsibility to make sure that in the input
CUBIT (or similar) mesh that this code will read in SPECFEM3D format
from files \texttt{nodes\_coords\_file} and \texttt{mesh\_file}
you have created layers of elements that constitute a layer of constant
thickness aligned with the coordinate grid axes (X, Y and/or Z), so
that this code can assign CPML flags to them. This code does NOT check
that (because it cannot, in any easy way). The mesh inside these CPML
layers does not need to be structured nor regular, any non-structured
mesh is fine as long as it has flat PML inner and outer faces, parallel
to the axes, and thus of a constant thickness. The thickness can be
different for the X, Y and Z sides. But for X it must not vary, for
Y it must not vary, and for Z it must not vary. If you do not know
the exact thickness, you can use a slightly LARGER value in this code
(say 2\% to 5\% more) and this code will fix that and will adjust
it; never use a SMALLER value otherwise this code will miss some CPML
elements.

@@

where \texttt{\textbf{domain\_ID}}\textbf{ is 1 for acoustic and 2
for elastic or viscoelastic materials,} \texttt{material\_ID} a unique
identifier, \texttt{rho} the density in $kg\, m^{-3}$, \texttt{vp}
the P-wave speed in $m\, s^{-1}$, \texttt{vs} the S-wave speed in
$m\, s^{-1}$, \texttt{Q} the quality factor and \texttt{anisotropy\_flag}
an identifier for anisotropic models. Note that the \texttt{Qkappa}
value is ignored by the code unless \texttt{FULL\_ATTENUATION\_SOLID}
is set. Note also that both \texttt{Qkappa} and \texttt{Qmu} are ignored
by the code unless \texttt{ATTENUATION} is set. If you want a model
with no \texttt{Qmu} attenuation, both set \texttt{ATTENUATION} to
\texttt{.false.} in the \texttt{Par\_file} and set \texttt{Qmu} to
9999 here. If you want a model with no \texttt{Qkappa} attenuation,
both set \texttt{FULL\_ATTENUATION\_SOLID} to \texttt{.false.} in
the \texttt{Par\_file} and set \texttt{Qkappa} to 9999 here.
Note that Qmu is always equal to Qs, but Qkappa is in general not equal to Qp. To convert one to the other see doc/note\_on\_Qkappa\_versus\_Qp.pdf and utils/attenuation/conversion\_from\_Qkappa\_Qmu\_to\_Qp\_Qs\_from\_Dahlen\_Tromp\_959\_960.f90.
